Implementation of Adaptive Waveform Inversion via a Penalty Function

### 2023-11-16 repro/papers/awigrad/paper.tex
interim

$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).

$[0,t_d]$ = acquisition time interval

$[-t_U,t_U]$ = adaptive filter time interval

$D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$ (again, no attempt to scale for geometry of acquisition).

$U$ = adaptive filter Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$

$w_* \in C_0^{\infty}(\bR)$ = known point source wavelet

$M = $ bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.

\begin{theorem}
  \label{thm:eu}
For $(\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying
\begin{itemize}
\item
  \begin{eqnarray}
    \label{eqn:awe}
    \frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
                                        w_*(t) \delta(\bx-\bx_s); \\
    \frac{\partial \bv}{\partial t} & = & - \beta \nabla p; \\
    p, \bv & = & 0 \mbox{ for }  t \ll 0.
  \end{eqnarray}
\item
  $p, \bv$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$.
\end{itemize}
\end{theorem}
\begin{proof}
  See Appendix A.
\end{proof}

Define $F: M \rightarrow D$ by
\[
  F[m](\bx_r,t;\bx_s) = p(\bx_r,t;\bx_s), \, (\bx_r,\bx_s) \in Q, \, t \in [0,t_D],
\]
for the pressure field $p$ solving \ref{eqn:awe}.

\begin{theorem}
  \label{thm:mbd}
  Under the hypotheses of Theorem \label{thm:eu},
  \begin{itemize}
  \item [1, ] $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
  \item [2. ] $F$ has a directional derivative $DF[m]: {\cal M} \rightarrow D$ for every $m \in M$.
    \item [3. ] for every $m \in M$, $DF[m]$ extends to a bounded map on ${\cal H} = L^2(\bR^3) \times L^2(\bR^3)$
  \end{itemize}
\end{theorem}
\begin{proof}
  See Appendix A.
\end{proof}

$F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.

@@

The convolution operator implemented in the module {\tt segyvc} actually implements a time-discrete version of $K$ (or $L$). Note that the transpose $K[u]^T]$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.

@@

Since operation is trace-by-trace, that is block-diagonal, $S[m] = \mbox{diag}_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$. Theorem \ref{thm:mbd} implies that $\|S[m]_{\bx_r,\bx_s}\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$. A similar statement applies to the derivative.

@@

According to Theorem \ref{thm:mbd} and the Observable Data Assumption, the left-hand side of equation \ref{eqn:blocknormal} is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$. According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense. Conclude that there exists $C>0$ so that
\begin{equation}
  \label{eqn:u0lower}
  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
\end{equation}
uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.

\subsection{AWI Penalty Operator}
The penalty operator is also block diagonal, that is, operates independently on each trace. For the trace at $(\bx_r,\bx_s) \in Q$, 
\begin{equation}
  \label{eqn:ppf}
  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
\end{equation}
That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$. The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained in \ref{eqn:HCSMWS:23a}
This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.

@@

Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of two least-squares problems, in sequence:

@@

@@

\subsection{Multiple Traces}
The general case follows immediately from the single trace case. It is just necessary to insert restriction to various (receiver, source) pairs in the relation \ref{eqn:singlegrad}, and sum. Note that the sum is implicit in the definition of $DF[m]^T$, and of $K^T$ and $S$, but must be carried out explicitly for the part of this expression that amounts to trace-by-trace scaling:
\[
  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d)\right.
\]
\[
  \left.  +\alpha^2 \sum_{(\bx_r,\bx_s)\in Q}\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\| u_{0,\sigma}[m](\bx_r,\cdot;\bx_s\|^{-2}\right.
\]
\[
  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d)_{(\bx_r,\bx_s)} \right.
\]
\begin{equation}
  \label{eqn:singlegrad}
\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) _{(\bx_r,\bx_s)} \right).
\end{equation}
This expression suggestions the following algorithm outline:
\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal};
\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 4;
\item construct linear operator $W_{\alpha,\sigma}[m]$ on $D$ that scales each trace at $(\bx_r,\bx_s) \in Q$  by $ \|tu_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-4}$
\item compute four vectors in $D$:
  \begin{enumerate}
  \item $G_1 = K[u_{\alpha,\sigma}[m]]^Tr_{\alpha}$
  \item $G_2 = K[v_0]^Tr_0$
  \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
  \item $G_0 = G_1 + \alpha^2W_{\alpha,\sigma}[m](G_2 + G_3)$
  \end{enumerate}
\item assemble
  $\nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^TG_0$
\end{enumerate}

>>>>

%$P = \bR^{|Q|}$ = real functions on $Q$, with Euclidean norm (no attempt to scale for geometry of acquisition).

The data to be inverted consists of $|Q|$ functions of time (``traces''). The time interval of recording is assumed to be the same for all traces, namely $[0,t_D]$. Traces are assumed square-integrable, members of $ L^2([0,t_D])$ - indeed the square of the trace $L^2$ norm is essentially the energy transferred from the surrounding acoustic fluid to the recording instrument \cite[]{SantosaSymes:00}. The set of traces forms the Hilbert space $D$ = data Hilbert space = $\bigoplus_{(\bx_r,\bx_s) \in Q} L^2([0,t_D])$. Note that no attempt is made to weight the traces for the spatial distribution of the acquisition coordinates.

The pressure field $p$, sampled at the receiver locations, is a component of the solution ${p,\bv}$ of the linear acoustic wave equation
\begin{eqnarray}
  \label{eqn:awe}
  \frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
                                      w_*(t) \delta(\bx-\bx_s); \\
  \frac{\partial \bv}{\partial t} & = & - \beta \nabla p; \\
  p, \bv & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
The right-hand side is localized in space at a source position $\bx_s \in Q_s$, with time dependence given by a {source wavelet} $w_* \in C_0^{\infty}(\bR)$. Details about the source mechanism are in fact difficult to measure directly and could justifiably be included in the parameters to be estimated in the solution of the inverse problem, but for the purposes of this discussion the wavelet $w_*$ is assumed known.

The acoustic parameter fields, bulk modulus ($\kappa$) and buoyancy ($\beta$) appearing in the system \ref{eqn:awe} are assumed smooth and uniformly bounded.
Since the speed of propagation is bounded as is the set of source locations $Q_s$ and the interval of propagation $[0,t_D]$, the values of $\kappa(\bx)$ and $\beta(\bx)$ for sufficiently lartge $|\bx|$ have no effect on the solution of the system \ref{eqn:awe}, so $\log \kappa$ and $\log \beta$ can be assumed square-integrable without loss of generality. The set of feasible models $M$ is a  bounded open subset of ${\cal M} = C^{\infty}(\bR^3) \times C^{\infty}(\bR^3) \cap L^{\infty}(\bR^3) \times L^{\infty}(\bR^3) \cap L^2(\bR^3) \times L^2(\bR^3)$.

Under these conditions, the following conclusions may be drawn about the system \ref{eqn:awe}:
\begin{itemize}
  \item[A1. ] For $m = (\log \kappa, \log \beta) \in M$, shot location ${\bf x}_s\in Q_s$, there exist distributions (pressure and velocity fields) $p(\cdot,\cdot;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ satisfying the system \ref{eqn:awe};
  \item[A2. ] $p(\cdot,\cdot,t;\bx_s)$, ${\bf v}(\cdot,\cdot;\bx_s)$ are smooth in the punctured space-time $\bR^4 \setminus \{(\bx_s,t): t \in \bR\}$;
    \item[A3. ] For $(\log \kappa, \log \beta) \in M$, define 
\[
  F[m](\bx_r,t;\bx_s) = p(\bx_r,t;\bx_s), \, (\bx_r,\bx_s) \in Q, \, t \in [0,t_D],
\]
for the pressure field $p$ solving \ref{eqn:awe}. The mapping ({\em forward map}, {\em modeling operator},...) $F: M \rightarrow D$ so defined satisfies
\begin{itemize}
  \item[A3.1 ] $\|F[m](\bx_r,\cdot;\bx_s)\|_{L^2([0,t_D])}$ is uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$.
  \item[A3.2 ] $F$ has a directional derivative $DF[m]: {\cal M} \rightarrow D$ for every $m \in M$;
  \item[A3.3 ] for every $m \in M$, $DF[m]$ extends to a bounded map on ${\cal H} = L^2(\bR^3) \times L^2(\bR^3)$.
  \end{itemize}
\end{itemize}
See \cite{Symes:23a} for detailed statements and justifications of these assertions.

{\bf Note:}
$F$ is the mapping implemented (in finite difference approximation) in {\tt asg.fsbop}, with the additional constraint that the buoyancy field is fixed so that the computational version of ${\cal M}$ consists of (gridded) bulk modulus fields.

Adaptive filters are collections of traces defined on a (possibly)
different time interval, usually conveniently chosen to be
time-symmetric, that is, $[-t_U,t_U]$. The trace geometry of adaptive
filters is otherwise the same as that of the data traces. $U$ will denote the  adaptive
filter Hilbert space: $U= \bigoplus_{(\bx_r,\bx_s) \in Q} L^2([-t_U,t_U])$.

@@

The convolution operator implemented in the module {\tt segyvc} actually implements a time-discrete version of $K$ (or $L$). Note that the transpose $K[u]^T$ is a restricted version of cross-correlation, implemented as the transpose of {\tt segyvc.ConvolutionOperator}.

@@

Since operation is trace-by-trace, that is block-diagonal, $S[m] = \mbox{diag}_{(\bx_r,\bx_s) \in Q}S[m]_{\bx_r,\bx_s}$. Fact A3.1 implies that $\|S[m]_{\bx_r,\bx_s}\|_{{\cal B}(L^2([-t_U,t_U]),L^2([0,t_D]))}$ is also uniformly bounded over $m \in M, (\bx_r,\bx_s) \in Q$. A similar statement applies to the derivative.

@@

According to Fact A3.1 and the Observable Data Assumption, the left-hand side of equation \ref{eqn:blocknormal} is bounded by a multiple of $\|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]}$, uniformly in $m \in M, (\bx_r,\bx_s) \in Q$. According to the Observable Data Assumption, the right-hand side is bounded below by a multiple of $\|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}$, uniform in the same sense. Conclude that there exists $C>0$ so that
\begin{equation}
  \label{eqn:u0lower}
  \|u_{0,\sigma}[m,d](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} \ge C \|d(\bx_r,\cdot;\bx_s)\|_{L^2[0,t_D]}
\end{equation}
uniformly in $m \in M, (\bx_r,\bx_s) \in Q$.

\subsection{AWI Penalty Operator}
The penalty operator is also block diagonal, that is, operates independently on each trace. For the trace at $(\bx_r,\bx_s) \in Q$, 
\begin{equation}
  \label{eqn:ppf}
  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
\end{equation}
That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$. The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.

@@

Therefore evaluating $\tilde{J}_{\alpha,\sigma}$ involves solution of
two least-squares problems, in sequence.

@@

To express the penalty operator $T_{\sigma}[m]$ conveniently, along
with similar operators appearing in the expression for the gradient
developed below, introduce functions
\begin{enumerate}
\item $R: U \rightarrow ({\bf R}^{+})^{|Q|}$,
  \[
    Ru((\bx_r,\bx_s)) = \|u(\bx_r, \cdot;\bx_s)\|
  \]
\item $W: {\bf R}^{|Q|} \times {\bf R} \times {\bf R}^{+}
  \rightarrow {\cal B}(D,D)$,
  \[
    W(q,p_s, p_t)f (\bx_r,t;\bx_s) = t^{p_t}
    q(\bx_r,\bx_s)^{p_s}f(\bx_r,t;\bx_s)
  \]
\end{enumerate}
Use the same notation for the similar function with $D$ replaced by
$U$. Note that $W$ is only well-defined if either $p_s \ge 0$ or $q>0$.

With these notations, $T_{\sigma}[m]$ may be expressed as
\begin{equation}
  \label{eqn:talt}
  T_{\sigma}[m] = W(Ru_{0,\sigma}[m],-1, 1)
\end{equation}

{\bf Note:} IWAVE implements the mappings $R$ and $W$ in the form of
the commands {\tt iwave/trace/main/rms.x} and {\tt
  iwave/trace/main/txgain.x}. The natural python implementation of $R$
would output the vector of trace norms as a python array (or a NumPy
ndarray). However that would require extracting the number of traces
from the input, and checking that the inputs are compatible. This is
already implicitly done in the IWAVE manipulations of SU trace
data. So {\tt iwave/trace/main/rms.x} stores the vector of trace norms
as the 0th data sample in a set of traces, one for each trace of the
input data: all header words are the same except for {\tt ns}, which
is set to 1.

The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
the two IWAVE commands, manipulating the data as SU data files, rather
than as VCL vectors. That is, these are simply functions, not {\tt
  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
have access to their data files (keyword {\tt data}), these is
convenient to use in the definitions of AWI penalty operator,
gradient, and so on.

@@

\subsection{Multiple Traces}
The general case follows immediately from the single trace case. It is
just necessary to observe that the scalare multiplication in the
expression \ref{eqn:singlegrad} is a trace-by-trace scaling that can
be computed by a suitable combination of the mappings $R$ and $W$,
defined above. The gradient is unlikely to be needed without the value
of $\tilde{J}_{\alpha,\sigma}$, so assume that
$T_{\sigma}[m]u_{\alpha,\sigma}$ is available, as well as the vector
of trace norms of $Ru_{0,\sigma}[m]$. Then multiplication of a trace
at $(\bx_r,\bx_s)$ by
$\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|
u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-2}$ is precisely application of
\[
  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)
\]
So
\[
  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d)\right.
\]
\[
  \left.  +\alpha^2 W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma},2,0)
( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) \right.
\]
\begin{equation}
  \label{eqn:singlegrad}
\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m])  \right).
\end{equation}
This expression suggestions the following algorithm outline:
\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute
  $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal}
  for $\alpha=0$, and $T_{\sigma}[m]=W(Ru_{0,\sigma},-1,1)$ (this has
  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$;
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by
  solving the normal equations \ref{eqn:normal}, and record
  $T_{\sigma}[m]u_{\alpha,\sigma}[m]$ as a by-product;
\item compute $v_{0\sigma} \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_{0,\sigma} = u_{0,\sigma}[m]$
\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 3;
\item compute four vectors in $D$:
  \begin{enumerate}
  \item $G_1 = K[u_{\alpha,\sigma}[m]]^Tr_{\alpha}$
  \item $G_2 = K[v_0]^Tr_0$
  \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
  \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
  \end{enumerate}
\item assemble
  $\nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^TG_0$
\end{enumerate}

\subsection{Warner-Guasch AWI}
The (``forward'') AWI objective defined in \cite{Warner:16} is
the limit
\[
  \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma})
\]
\begin{equation}
  \label{eqn:wg}
  = \frac{1}{2} \|T_{\sigma}[m]u_{0,\sigma}\|^2
\end{equation}
\cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is
manifestly not the case for $\sigma > 0$, however for small $\sigma$
the error is presumably negligible. Neglecting it, the gradient becomes
\[
  \nabla \tilde{J}_{\sigma}[m] = DF[m]^T
    W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{0,\sigma}[m],2,0)
\]
\begin{equation}
  \label{eqn:wggrad}
\times K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]).
\end{equation}
